A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-03-03 response/response.tex

    \item
        In lines 47-49, you say “The speciﬁc parameters ... can be found ... in the
        paper repository.” It seems a good idea to have the paper repository with ﬁles that a
        user can eventually use. However, the content of the paper should be independent
        of the repository. Please include a table with the parameters used. Similarly, in
        page 11 (lines 11-13).

{\color{red}  
\begin{itemize}
    \item
        Correction made in the manuscript. 
  \end{itemize}}

  \item
    Pg. 14: In line 55, is there any conclusion you can draw from the power law index
    $1.0 * 10^-6$?

{\color{red}  
\begin{itemize}
     \item
      Manuscript updated. More significant than the power law index is that the methods have the same power law relationship with computation time at all, meaning that there is some weak evidence that the methods have the same complexity class. 
  \end{itemize}}

### 2014-03-03 jacobian_calc_CMAME.tex
changes denoted more plainly

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method.  The distinction of ``accurate'' is
defined by comparison of the \note[Michael]{re: I.1}\remove[Michael]{new}
method to popular finite differencing techniques used for computing
tangent-stiffness matrices and with the exact to machine precision
algorithmically consistent derivatives computed with \emph{automatic
differentiation}. For the reader's convenience, from plasticity theory, the algorithmic consistency of
the tangent-stiffness matrix refers to consistency with the integration of a
plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is
computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.

@@

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing complex-step in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. This open source resource is available at the corresponding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
 or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

@@

\subsubsection{The ``complex-step'' method.}  \label{sec:CSmethod} It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left (
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im}
\left( f (x + i h) \right)}{h} +\mathcal{O}\left ( h^2 \right).  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite-differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function.

For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left ( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left ( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
    Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

\note[Michael]{re: I.4}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an infinite series form and the computing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

@@

A series of nine single compute core (i.e. serial) test problems were run with
the discretization size being refined with every test in the series. The aim of
increasing the refinement was to examine differences between the methods in
terms of accuracy and speed. This series would comprise the single core runs.
The specific parameters of these tests can be found in \emph{Peridigm} input
{\tt xml} files included in the paper repository, however the parameters
are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow
users with the appropriately modified version of \emph{Peridigm} to reproduce
the results found in this paper.

\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Test Parameters} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon & Load rate & # Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\meter}& \si{\centimeter\per\second} &   &     \\
        \midrule
        1000 & 1.99             & .3   & 2.0 & 2 & 10^-8\\
        2000 & 4.25             & .24  & 2.0 & 2 & 10^-8\\
        3000 & 7.79             & .21  & 2.0 & 2 & 10^-8\\
        4000 & 15.4             & .19  & 2.0 & 2 & 10^-8\\
        6000 & 20.0             & .16  & 2.0 & 2 & 10^-8\\
        8000 & 31.4             & .15  & 2.0 & 2 & 10^-8\\
        12000 & 40.1            & .131 & 2.0 & 2 & 10^-8\\
        16000 & 83.9            & .12  & 2.0 & 2 & 10^-8\\
        32000 & 165.0           & .095 & 2.0 & 2 & 10^-8\\
        1000000 & 1670          & .03  & 2.0 & 2 & 10^-8\\
        \bottomrule
    \end{tabular}
\end{table*}

@@

\section{Results and Discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived Eqn.~(\ref{eqn:accuracy}) and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$& $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$& $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

\subsection{Speed Data} Speed measurements were taken according to the methods
mentioned in Section~\ref{tcs}. The averaged results of these speed
measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.
Average time in seconds to compute a TS matrix is plotted as a function of the
number of nonzero TS elements.  CS was the slowest of the four methods, taking
the longest average time per iteration,  followed by CD, FD and AD in that
order. Calculation time has a power-law relationship with problem size, with a
power-law index of roughly $\num{1.0E-6}$ for each of the methods. This bodes well
for the scalability of each of the methods by themselves in \emph{Peridigm}. If we
assumed every individual operation, generally speaking, took the same amount of
time on average such that compute time were a measure of number of operations,
this suggests that the methods as in \emph{Peridigm} have the same order
algorithm complexity when considering computing a TS matrix alone, that is
computation time is $\mathcal{O}(10^{(10^{-6} \times \log_{10}n)})$. This would
suggest, to the algorithm design way of thinking, that the methods are equally
well suited to the task of computing a TS matrix, if it were further assumed
that the TS matrices produced by each method were identical. This is not the case
as seen in the following subsection.
%
\begin{figure}[tbp] \centering \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
\caption{Serial test series speed measurements.} \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in
Fig.~\ref{fig:multi_speed}. For this series, rather than varying the number of
nonzero TS matrix elements, the number of compute cores used to solve the
problem is increased, while the number of nonzero TS matrix elements remained
constant at $\num{1.67E9}$ .  CS again was the slowest method, followed by CD,
AD and then FD.  There is a somewhat surprise reversal of FD and AD for these
parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking
at average iteration time, the factor of speed improvement scales inversely with
the number of compute cores used for the given conditions of this group of
tests. Looking at Fig.~\ref{fig:multi_speed}, the reader can see that when core
count doubles, compute time halves, for each method. Again, these results show
that each of the methods as implemented in \emph{Peridigm} scale well when
additional processors are used. 
%
\begin{figure}[tbp] \centering \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
\caption{Multicore test series speed measurements.} \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed
tests with respect to the serial results is that the AD data-structures carry a
list of partial derivatives of the functional evaluations along side the
functional evaluations themselves only to be processed at the end of the
tangent-stiffness calculation when the derivatives are requested. This
represents additional load on the hardware memory subsystem, such that may
exceed that of FD yet still fall under a supposed memory bandwidth threshold
during serial operation. Cores within a multicore processor use the same shared
bus to access memory, such that cores compete for memory access. Supposing that
exceeding the memory bandwidth threshold penalizes additional requested
bandwidth, it is likely that the decrease in available memory bandwidth per
core caused by parallel operation coupled with AD's additional resource
requirements causes the AD method to be slower than FD.

@@

Data for accuracy measurements were taken according to the methods mentioned in
Section~\ref{tcs}, and reduced using Eqn.~(\ref{eqn:accuracy}). The
averaged results of these accuracy measurements for the serial tests can be
seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as
$l^2$-norm in the figure labels for conciseness) of the element-wise difference
between a given TS matrix and the TS matrix produced by the AD method is
plotted as a function of the number of nonzero TS matrix elements. CS was shown
to be the most accurate of the methods when compared to AD and was roughly
\emph{$6$ orders of magnitude more accurate than CD} and \emph{$9$ orders of
magnitude more accurate than FD}. This accuracy ranking order was maintained
for each test in the series. It should be pointed out that compared to the
$l^2$-norms of any of the TS matrices by themselves, the magnitude of the
accuracy metric calculated with Eqn.~\ref{eqn:accuracy} for any of the methods
is relatively small. Relative error was computed from these measurements and is reported in
Table~\ref{tab:RelAcc}. The values for the $l^2$-norms of the TS matrices appear
in the simulation data which is available in the paper repository. The effect
of these accuracy differences on convergence rate of the Newton solver was studied and the
results are reported in section~\ref{sec:PeridigmConvergenceStudy}.  
%
The accuracy of FD and CD could be improved by using a smaller probe distance,
$h$; however, if $h$ is too small, then the methods can suffer from severe
effects, including solution divergence, of the round-off error associated with
subtracting two numbers that are very close to one another.  The value of $h$
used in this study is the default value used in \emph{Peridigm} which is
heuristically derived based on computational node spacing using the method explained in \cite[pp. 90]{ref-Adaggio}.  
\add[Michael]{To summarize, a spacing of $1.E-6$ times grid spacing is used to provide acceptable accuracy but prevent roundoff error, 
as is not a user determined setting in \emph{Peridigm}.} One of the beauties
of the CS method is that since dependence on $h$ decreases for small $h$, which
a numerical example in \cite[Table 1]{squire1998using} demonstrates, $h$,
cannot be too small, therefore it can be set to machine epsilon and the
analysis performed without the issues associated with round-off error. The
value chosen for $h$ for CS used in this study was $\num{1.E-100}$ to
demonstrate this.
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/serial_accuracy.pgf}} \caption{Serial test series
accuracy measurements.} \label{fig:serial_accuracy} \end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test
series can be seen in Fig.~\ref{fig:multi_accuracy}. The ordinate axis units
remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the abscissa now
indicates number of compute cores used to solve the test problem. Accuracy
trends matched those for the single core tests. 
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/multi_accuracy.pgf}} \caption{Multicore test
series accuracy measurements.} \label{fig:multi_accuracy} \end{figure}
%
It can be seen in Figs.~\ref{fig:serial_accuracy} and ~\ref{fig:multi_accuracy} that there appears to be some fluctuating behavior in the accuracy of the methods. Those figures are depicting absolute error. If relative error is computed from the absolute error measurements by dividing them by the Frobenius norm of the appropriate AD produced TS matrix, it can be seen as in Table~\ref{tab:RelAcc}, that there is little variation in the accuracy of the methods between tests or between serial and parallel test groups.

>>>>

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method.  The distinction of ``accurate'' is
defined by comparison of the \note[Michael]{re: I.1}\remove[Michael]{new}
method to popular finite differencing techniques used for computing
tangent-stiffness matrices and with the exact to machine precision
algorithmically consistent derivatives computed with \emph{automatic
differentiation}. \note[Michael]{re: II.1}For the reader's convenience, 
from plasticity theory, the algorithmic consistency of
the tangent-stiffness matrix refers to consistency with the integration of a
plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is
computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.

@@

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing complex-step in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. \note[Michael]{re: II.5}This open source resource is available at the corresponding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

@@

\subsubsection{The ``complex-step'' method.}  \label{sec:CSmethod} It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left (
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im}
\left( f (x + i h) \right)}{h} +\mathcal{O}\left ( h^2 \right).  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite-differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function. 
%
\note[Michael]{re: II.6}
For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left ( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left ( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

\note[Michael]{re: I.4, John all the rest of this section is new}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an infinite series form and the computing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

@@

A series of nine single compute core (i.e. serial) test problems were run with
the discretization size being refined with every test in the series. The aim of
increasing the refinement was to examine differences between the methods in
terms of accuracy and speed. This series would comprise the single core runs. 
\note[Michael]{re: II.16, specific test parameters were asked for}
The specific parameters of these tests can be found in \emph{Peridigm} input
{\tt xml} files included in the paper repository, however the parameters
are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow
users with the appropriately modified version of \emph{Peridigm} to reproduce
the results found in this paper.

\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Test Parameters} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon & Load rate & # Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\meter}& \si{\centimeter\per\second} &   &     \\
        \midrule
        1000 & $1.99$             & $.3$   & $2.0$ & $2$ & $\num{10E-8}$ \\
        2000 & $4.25$             & $.24$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        3000 & $7.79$             & $.21$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        4000 & $15.4$             & $.19$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        6000 & $20.0$             & $.16$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        8000 & $31.4$             & $.15$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        12000 & $40.1$            & $.131$ & $2.0$ & $2$ & $\num{10E-8}$ \\
        16000 & $83.9$            & $.12$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        32000 & $165.0$           & $.095$ & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

\section{Results and Discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived Eqn.~(\ref{eqn:accuracy}) and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

\subsection{Speed Data} Speed measurements were taken according to the methods
mentioned in Section~\ref{tcs}. The averaged results of these speed
measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.
Average time in seconds to compute a TS matrix is plotted as a function of the
number of nonzero TS elements.  CS was the slowest of the four methods, taking
the longest average time per iteration,  followed by CD, FD and AD in that
order. Calculation time has a power-law relationship with problem size, with a
power-law index of roughly $\num{1.0E-6}$ for each of the methods. This bodes well
for the scalability of each of the methods by themselves in \emph{Peridigm}. If we
assumed every individual operation, generally speaking, took the same amount of
time on average such that compute time were a measure of number of operations,
this suggests that the methods as in \emph{Peridigm} have the same complexity class
when considering computing a TS matrix alone, that is
computation time is $\mathcal{O}(10^{(10^{-6} \times \log_{10}n)})$. The definition of 
complexity class used here is the informal one given in \cite[pp. 1059]{cormen2009introduction}. This would
suggest that the methods are equally well suited to the task of computing a TS matrix, if it were further assumed
that the TS matrices produced by each method were identical. This is not the case
as seen in the following subsection.
%
\begin{figure}[tbp] \centering \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
\caption{Serial test series speed measurements.} \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in
Fig.~\ref{fig:multi_speed}. For this series, rather than varying the number of
nonzero TS matrix elements, the number of compute cores used to solve the
problem is increased, while the number of nonzero TS matrix elements remained
constant at $\num{1.67E9}$ .  CS again was the slowest method, followed by CD,
AD and then FD.  There is a somewhat surprise reversal of FD and AD for these
parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking
at average iteration time, the factor of speed improvement scales inversely with
the number of compute cores used for the given conditions of this group of
tests. Looking at Fig.~\ref{fig:multi_speed}, the reader can see that when core
count doubles, compute time halves, for each method. Again, these results show
that each of the methods as implemented in \emph{Peridigm} scale well when
additional processors are used. 
%
\begin{figure}[tbp] \centering \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
\caption{Multicore test series speed measurements.} \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed
tests with respect to the serial results is that the AD data-structures carry a
list of partial derivatives of the functional evaluations along side the
functional evaluations themselves only to be processed at the end of the
tangent-stiffness calculation when the derivatives are requested. This
represents additional load on the hardware memory subsystem, such that may
exceed that of FD yet still fall under a supposed memory bandwidth threshold
during serial operation. Cores within a multicore processor use the same shared
bus to access memory, such that cores compete for memory access. Supposing that
exceeding the memory bandwidth threshold penalizes additional requested
bandwidth, it is likely that the decrease in available memory bandwidth per
core caused by parallel operation coupled with AD's additional resource
requirements causes the AD method to be slower than FD.

@@

Data for accuracy measurements were taken according to the methods mentioned in
Section~\ref{tcs}, and reduced using Eqn.~(\ref{eqn:accuracy}). The
averaged results of these accuracy measurements for the serial tests can be
seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as
$l^2$-norm in the figure labels for conciseness) of the element-wise difference
between a given TS matrix and the TS matrix produced by the AD method is
plotted as a function of the number of nonzero TS matrix elements. CS was shown
to be the most accurate of the methods when compared to AD and was roughly
\emph{$6$ orders of magnitude more accurate than CD} and \emph{$9$ orders of
magnitude more accurate than FD}. This accuracy ranking order was maintained
for each test in the series. It should be pointed out that compared to the
$l^2$-norms of any of the TS matrices by themselves, the magnitude of the
accuracy metric calculated with Eqn.~\ref{eqn:accuracy} for any of the methods
is relatively small. Relative error was computed from these measurements and is reported in
Table~\ref{tab:RelAcc}. The values for the $l^2$-norms of the TS matrices appear
in the simulation data which is available in the paper repository. The effect
of these accuracy differences on convergence rate of the Newton solver was studied and the
results are reported in section~\ref{sec:PeridigmConvergenceStudy}.  
%
The accuracy of FD and CD could be improved by using a smaller probe distance,
$h$; however, if $h$ is too small, then the methods can suffer from severe
effects, including solution divergence, of the round-off error associated with
subtracting two numbers that are very close to one another.  The value of $h$
used in this study is the default value used in \emph{Peridigm} which is
heuristically derived based on computational node spacing using the method explained in \cite[pp. 90]{ref-Adaggio}.  
\add[Michael]{To summarize, a spacing of $1.E-6$ times grid spacing is used to provide acceptable accuracy but prevent roundoff error, 
as is not a user determined setting in \emph{Peridigm}.} One of the beauties
of the CS method is that since dependence on $h$ decreases for small $h$, which
a numerical example in \cite[Table 1]{squire1998using} demonstrates, $h$,
cannot be too small, therefore it can be set to machine epsilon and the
analysis performed without the issues associated with round-off error. The
value chosen for $h$ for CS used in this study was $\num{1.E-100}$ to
demonstrate this.
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/serial_accuracy.pgf}} \caption{Serial test series
accuracy measurements.} \label{fig:serial_accuracy} \end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test
series can be seen in Fig.~\ref{fig:multi_accuracy}. The ordinate axis units
remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the abscissa now
indicates number of compute cores used to solve the test problem. Accuracy
trends matched those for the single core tests. 
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/multi_accuracy.pgf}} \caption{Multicore test
series accuracy measurements.} \label{fig:multi_accuracy} \end{figure}
%
\note[Michael]{re: II.22, relative accuracy figures were asked for, these were taken from the archive data}
It can be seen in Figs.~\ref{fig:serial_accuracy} and ~\ref{fig:multi_accuracy} that there appears to be some fluctuating behavior in the accuracy of the methods. Those figures are depicting absolute error. If relative error is computed from the absolute error measurements by dividing them by the Frobenius norm of the appropriate AD produced TS matrix, it can be seen as in Table~\ref{tab:RelAcc}, that there is little variation in the accuracy of the methods between tests or between serial and parallel test groups.
